---
title: "MMAPPR2 workflow"
author: "Kyle Johnsen, Nathaniel Jenkins, and Jonathon Hill"
output: html_document
vignette: >
  %\VignetteIndexEntry{MMAPPR2 workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Scope

This vignette describes the current MMAPPR2 workflow and its public analysis controls.
MMAPPR2 uses Bioconductor APIs for BAM/FASTA access and local coding-effect annotation;
external GMAP, command-line `samtools mpileup`, Ensembl VEP, `species`, and `VEPFlags`
are not required by the current workflow.

MMAPPR2 maps a phenotype from pooled F2 RNA-seq in two conceptual phases:

1. compare mutant and phenotypically WT pools at covered A/C/G/T positions to locate a
   linkage region;
2. inspect mutant sequence changes inside the refined region, attach WT-pool evidence,
   and annotate coding consequences.

# Inputs

The constructor expects:

- one or more WT-pool BAM files;
- one or more mutant-pool BAM files;
- the reference FASTA used for alignment;
- a GTF annotation corresponding to that reference.

BAM and FASTA indexes are created when missing. The input GTF is read through
Bioconductor and is **not** destructively uncompressed or shell-processed.

```{r setup, eval=FALSE}
param <- mmapprParam(
  wtFiles = c("wt_pool.bam"),
  mutFiles = c("mutant_pool.bam"),
  refFasta = "reference.fa",
  gtf = "annotation.gtf",
  outputFolder = "mmappr2_results"
)
```

# Important defaults and explicit controls

```{r parameters, eval=FALSE}
param <- mmapprParam(
  wtFiles = "wt_pool.bam",
  mutFiles = "mutant_pool.bam",
  refFasta = "reference.fa",
  gtf = "annotation.gtf",
  minDepth = 20,
  minBaseQuality = 20,
  minMapQuality = 20,
  maxPileupDepth = 1000,
  candidateMinDepth = 1,
  candidateMinAltDepth = 2,
  candidateMinAltFreq = 0.80,
  # These WT-specific hard filters are optional by default. WT evidence is
  # still reported for every candidate but does not change the default candidate ordering.
  candidateMaxWtAltFreq = 1.00,
  candidateMinDeltaAF = 0.00,
  # Keep the fast one-shot pileup when it fits; otherwise chunk adaptively.
  candidatePoolMode = "auto",
  candidateChunkSize = 0,
  peakCutoffSd = 3,
  # Default: use the compatibility cutoff; "global_sd" selects the genome-wide-SD alternative.
  peakCutoffMethod = "legacy_current",
  peakIntervalWidth = 0.80,
  # Conservative default: span all selected high-density modes.
  peakIntervalMethod = "hpd_span",
  peakResampleIterations = 1000,
  randomSeed = 1,
  pairedEnd = FALSE,
  ignoreStrand = FALSE,
  expressionPseudocount = 0.01,
  # Export the evaluated AICc span curve for each fitted chromosome.
  exportAiccPlots = FALSE
)
```


Candidate pileup uses `candidatePoolMode="auto"` by default. The auto mode keeps the
fast single-pileup path when the refined interval is comfortably sized for available
memory, including container/cgroup limits on Linux, and switches to bounded genomic
chunks only when useful. In bounded-memory mode, BAMs are pooled incrementally so all
per-BAM pileup tables do not need to coexist in memory; WT support for candidate loci
uses the same strategy. `candidateChunkSize=0` selects an adaptive chunk width; a
positive value supplies an explicit maximum chunk width. `candidatePoolMode="memory"`
prefers a one-shot pileup, while `"chunked"` always uses bounded chunks. If an actual
allocation failure occurs on the fast mutant path, MMAPPR2 retries with smaller chunks
and uses temporary RDS files under R's `tempdir()` for already-filtered chunk results.
Those files are uniquely named and removed automatically.

When `exportAiccPlots=TRUE`, `outputMmapprData()` writes `aicc_plots.pdf` from the AICc
evaluations already stored by `loessFit()`; the export step does not rerun model fitting.

The two peak-method parameters are intentionally explicit because changing them changes
the statistical interpretation of a run. `peakCutoffMethod="legacy_current"` preserves
the compatibility cutoff calculation; `"global_sd"` instead uses the genome-wide
median of fitted values plus `peakCutoffSd` times their ordinary standard deviation.
Likewise, `peakIntervalMethod="hpd_span"` spans all KDE grid locations selected by the
high-density mass rule, whereas `"shortest_contiguous"` chooses one shortest contiguous
interval and can intentionally exclude a secondary mode.

For paired-end RNA-seq, set `pairedEnd=TRUE` so the descriptive expression summary
counts fragments appropriately. `ignoreStrand=FALSE` uses strand-aware
`summarizeOverlaps()` counting. The descriptive expression log2 fold change adds a
small pseudocount of 0.01 to both pool means by default, preventing Inf/NaN values
when a mean count is zero. Set `expressionPseudocount=0` to request the raw ratio.
This output is **not** a formal DESeq2/edgeR differential-expression test.

# Run the pipeline

```{r run, eval=FALSE}
md <- mmappr(param)
```

For learning or debugging, run the stages separately:

```{r stages, eval=FALSE}
md <- mmapprData(param)
md <- calculateDistance(md)  # pooled A/C/G/T frequencies and ED^p
md <- loessFit(md)           # chromosome-wise robust LOESS
md <- prePeak(md)            # genome-wide peak threshold
md <- peakRefinement(md)     # reproducible half-marker resampling
md <- generateCandidates(md) # mutant SNVs + WT evidence + coding effects
outputMmapprData(md)
```

# Candidate-stage output

The current lightweight candidate caller is still SNV-focused. For each candidate it
records mutant depth/allele frequency and WT depth/allele frequency. A variant whose
ALT allele is high-frequency in both pools is marked `sharedBackgroundLike` as reporting
metadata. By default this flag does **not** change the candidate order.
If either WT-specific hard filter is enabled, candidates without finite WT evidence for
that filter do not pass it; the permissive defaults avoid that low-coverage exclusion.

MMAPPR2 writes:

- `AllCandidateVariantsFor<seq>.tsv`: all retained SNV candidates, including noncoding
  candidates and WT-vs-mutant evidence;
- `DetectedMutationsFor<seq>.tsv`: coding consequences from `predictCoding()` (established filename preserved);
- `DifferentiallyExpressedGenesFor<seq>.tsv`: the historical filename for the descriptive raw-count fold-change screen;
- `genome_plots.pdf`, `peak_plots.pdf`, `mmappr2.log`, and `mmappr_data.RDS`;
- `aicc_plots.pdf` when `exportAiccPlots=TRUE` and finite AICc search results are available.

# Current methodological limits

Some analyses require a larger, separately validated methodological extension. In particular:

- exact indel/haplotype-aware candidate calling is not implemented;
- the peak data model still represents one dominant refined locus per chromosome;
- the >80% mutant-ALT default still reflects a simple recessive/high-fixation model,
  although it is now configurable;
- RNA-seq still has uneven genic coverage, allele-specific-expression effects, and
  reference-alignment bias;
- noncoding and structural effects are retained as sequence candidates where observable,
  but only coding SNV effects receive `predictCoding()` annotations;
- the expression summary is descriptive rather than inferential.

